Nail: A practical interface generator for data formats

### 2014-04-30 paper/osdi14/design.tex
Design

\subsection{Overview}

Nail grammars describe both the \emph{external format} and an \emph{internal representation} of a
protocol. Typical grammar languages describe only the former and leave it up to the programmer to
construct an internal representation, typically in the form of an abstract syntax tree.

Nail produces the following from a single, descriptive grammmar:

\begin{itemize}
\item type declarations for the internal model,
\item the \textit{parser}, a function to parse a sequence of bytes into an
instance of the model, and
\item the \textit{generator}, a function to create a
sequence of bytes from an instance of the model.
\end{itemize}

\noindent
In the rest of this section, we present the design of the Nail language.

\subsection{Basics}
Nail grammars are a set of rules. Each rule assigns a parser, which can be a reference to another
rule, a trivial parser describing an integer of a certain width or a combinator that combines
multiple parsers in some way. We provide the most common combinators familiar from other parser
combinator libraries, such as Parsec\cite{LeijenMeijer:parsec} and Hammer\cite{hammer-parser}.
However, we extend the sequence and choice combinators normally found in parser combinator libraries
with \emph{semantic labels} that name fields in the internal models. The syntax described by a Nail
grammar does not depend on these semantic labels.

@@

Dependent fields are defined within a structure like normal fields, but their name starts with an \@-sign. 
A dependent field is in scope and can be referred to by the definition of all subsequent fields in the same structure.

@@

% Data formats often contain redundant information, such as checksums or duplicated values.
% Conventional parsers expose all this information and validate it manually, which can lead to bugs.

% Another problem for interface generators is that binary protocols often contain length and offset
% fields. Conventional parsing algorithms can, in principle, deal with bounded offset fields - a finite
% automaton can count a bounded integer, by introducing a state for each value. However, such grammars
% are very large, making the state-machine implementation inefficient and the grammar very complicated.

% Nail will properly support both offset and length fields and much of the
% following discussion applies to both, although the current prototype only
% implements lengths, which we will focus on.

% We call length or offset fields \textit{dependent fields}, because during
% parsing, another parser depends on them, and while generating output, their
% value depends on some other structure in the data model. Dependency fields
% appear in a structure combinator as would any other integer field, but
% their name begins with an \texttt{@} sign.  A dependency field has to
% appear in the grammar before it can be used.

% Dependency fields are not exposed in the data model, but instead are
% transparently computed.  This frees the developer from checking that
% these fields are correct (for input) or having to keep their values in
% sync with the rest of the data structure (for output).

% \paragraph{Length fields.}

% The length combinator, \texttt{n_of}, takes a dependency field $n$
% and a parser, evaluates the parser exactly $n$ times (i.e., setting
% the number of iterations to be the $n$ field's value), and returns
% an array of the parser's values. When generating output, it emits
% the array and writes its length to the dependency field.

% \paragraph{Offsets.}

% The \texttt{offset} combinator takes a dependency field and a parser. It
% moves the parser to the position specified in the offset field and
% invokes the inner parser, and then moves the input back to its original
% position.  While generating output, all structures referred to by offset
% are generated after the main structure and the dependent offset fields
% are patched up.

% \paragraph{Checksums.}

% In many data formats, some values depend on external representation,
% such as checksums and cryptographic signatures.  While it would be
% possible to extend our constraint language to be powerful enough to
% support such constructs, we would essentially be building a separate,
% Turing-complete language that has all the same pitfalls existing programs
% have.  Therefore, we intend to allow the programmer to carefully escape
% Nail's programming model and write a function that takes a pointer to
% the dependent value and a range of bytes, using the \texttt{raw_depend}
% combinator.

>>>>

\subsection{Goals}
\paragraph{Internal Model.}
Nail grammars describe both the \emph{external format} and an \emph{internal representation} of a
protocol, as opposed to existing grammar languages describe only the former and leave it up to the programmer to
construct an internal representation, typically in the form of an abstract syntax tree.

Nail produces the following from a single, descriptive grammmar: 
\begin{itemize}
\item type declarations for the internal model,
\item the \textit{parser}, a function to parse a sequence of bytes into an
instance of the model, and
\item the \textit{generator}, a function to create a
sequence of bytes from an instance of the model.
\end{itemize}

\noindent

\paragraph{Semantic Bijecction.}
Second, this well-defined internal representation allows Nail to invert the
parser and generate output, while maintaining a \emph{semantic bijection}, that is, when generating
the corresponding output to a parser input, the two byte streams will have the same meaning, i.e. be
interpreted the same by Nail, but might not be identical.
A bijection in the traditional sense often does not make sense when parsing data formats. 
Consider a grammar for a text language that tolerates white space, or a binary protocol that tolerates
arbitrarily long padding.\footnote{Say, the physical layer of most communication
  protocols is a possibly infinite sequence of symbols that are syntactically
  nil followed by a pre-determined synchronization sequence and the actual
  contents of the transmission.} Program semantics should be
independent of the number of padding elements in the input, and Nail therefore
does not expose that information to the programmer. We call such discarded
fields \emph{constants}. 
Similarly, Nail does not preserve the layout of
objects referred to by their offsets. If the grammar consists of a simple protocol without offset
fields, constants and the like, there is a normal bijection between the internal model and valid
parser inputs.
\paragraph{Hide redundant information.}
Nail's internal model is designed to hide unneeded information from the application. The more
information Nail automatically deduces, the less the application can produce invalid data.
Therefore, Nail introduces \emph{dependent fields}, which contain data that can be computed during
generation and needs to be kept as additional state during parsing.

\paragraph{Allow parser extension.}
Real-world protocols contain complicated ways of encoding data. Representing these in an
intentionally limited model such as our parser language is complicated. Therefore, Nail introduces
\emph{transformations}, which allow arbitrary code by the programmer to interact with the parser and
generator. Nail parsers and generators interact with data through an abstract stream, which allows
reading and writing of bits as re-positioning. Transformations allow the programmer to write
functions in a general purpose language that consume streams and define new temporary streams, while
also reading or writing the value of dependent fields.

Initial versions of Nail included an offset combinator, which consumed a dependent field and applied
a parser at the offset specified therein. However, it proved impossible to foresee all the ways in which a protocol could encode an offset, for example
some protocols such as PDF and ZIP locate structures by scanning for a magic number starting at the
end of the file or a fixed offset. In nested grammars, offsets are also not necessarily from
the beginning of a file or packet.

We will introduce transformations, which allow the programmer to write arbitrary functions that can
handle such structures and streams, which are a generic abstraction for input and output data that
allow the decoded data to be integrated with the rest of the generated Nail parser.

\XXX[In the rest of this section, we first present the design of the Nail language and then show how
Nail's features match these goals.]
\subsection{Basics}
Nail grammars are a set of rules. Each rule assigns a parser, which can be a reference to another
rule, a trivial parser describing an integer of a certain width or a combinator that combines
multiple parsers in some way. We provide the most common combinators familiar from other parser
combinator libraries, such as Parsec\cite{LeijenMeijer:parsec} and Hammer\cite{hammer-parser}.
However, we extend the sequence and choice combinators normally found in parser combinator libraries
with \emph{semantic labels} that name fields in the internal models. The syntax described by a Nail
grammar does not depend on these semantic labels.

@@

Dependent fields are defined within a structure like normal fields, but their name starts with an \@-sign. 
A dependent field is in scope and can be referred to by the definition of all subsequent fields in
the same structure. Dependency fields can be passed to rule invocations as parameters.

@@

In order to handle complicated input functions, Nail allows the programmer to
break out of Nail's generated code in a controlled fashion, by manipulating dependent fields and
streams. 
By default, Nail implements streams with in-memory buffers, but the programmer can override this implementation.
While parsing or generating a Nail grammar, there is always an implicit \emph{current stream} and
referred to as \texttt{\$current} in the Nail language.

The programmer can define additional temporary streams by writing a \emph{transformation}, a pair of
functions, one for parsing and one for generating output. Temporary streams follow the naming and
scoping convention of dependent fields, but their names start with a \$-sign.

The parsing function of the transform takes any number of streams, either temporary or the current,
and outputs any number of temporary streams. The parsing function may reposition and read from the
input streams and read the values of dependent fields, but not change their values. The generating
function has to be an inverse of the parsing function. It takes the same set of temporary streams
the parsing function returned and is supposed to change the values of the same set of values the
generating function produced.

The programmer has to guarantee memory safety and unambiguity of the stream transformations. To
reduce both programmer effort and the risk of unsafe operations, we provide implementations of
transformations for many common features, such as checksums and size and offset fields. Furthermore,
we provide library functions that can be used to safely operate on streams, such as splitting and
concatenation. Nail implements streams as iterators, so they can share underlying buffers and can be
efficiently duplicated and split.

Transformations can handle a wide variety of patterns in data formats, including the following: 
\paragraph{Offsets.}
We provide a transform for handling offset fields that is invoked as follows:
\texttt{$fragment transform offset\_u32($file_stream, @offset)}. 
They are implemented by two  functions:
\begin{figure}[tb]
\begin{verbatim}
int offset_u32_parse(NailStream *out_fragment,NailStream *in_current,const uint32_t *off){
   out_fragment = suffix of in_current at offset off
}
int offset_u32_generate(NailStream *in_fragment, NailStream *out_current,uint32_t *off){
   off = position of out_current
   append in_fragment to out_current
}
\end{verbatim}
\caption{Pseudocode for the two C functions implementing an offset transformation.}
\end{figure}

\paragraph{Sizes.}
A similar transform handles size fields. Just like the offset transform, it takes two parameters, a
stream and a dependent field, but instead of returning the suffix of the current stream after an
offset, it returns a slice of the given size from the current stream starting at its current
position. When generating, it appends the fragment stream to the current stream and writes the size
of the fragment to the dependent field.

\paragraph{Compressed data.}
Transforms can handle encoded or compressed data. \XXX[More]
\paragraph{Checksum}
Checksums can also be implemented using transform \XXX[How?]

\noindent A real-world example with many different transforms is explained in Section~\ref{s:eval-format-zip}.

### 2014-04-30 paper/osdi14/intro.tex
Design

Second, this well-defined internal representation allows Nail to invert the
parser and generate output, while maintaining a \emph{semantic bijection}.
A bijection in the traditional sense often does not make sense when parsing data formats. 
Consider a grammar for a
text language that tolerates white space, or a binary protocol that tolerates
arbitrarily long padding.\footnote{Say, the physical layer of most communication
  protocols is a possibly infinite sequence of symbols that are syntactically
  nil followed by a pre-determined synchronization sequence and the actual
  contents of the transmission.} Program semantics should be
independent of the number of padding elements in the input, and Nail therefore
does not expose that information to the programmer. We call such discarded
fields \emph{constants}. 
Similarly, Nail does not preserve the layout of
objects referred to by their offsets. If the grammar consists of a simple protocol without offset
fields, constants and the like, there is a normal bijection between the internal model and valid
parser inputs.

@@

Third, Nail introduces two abstractions, \emph{dependent fields} and \emph{structural transformations}, to
elegantly handle problematic features of data formats. Dependent fields are fields in the protocol
whose value depends in some way on the value or layout of other parts of the protocol, such as
length, offset and checksum fields. Structural transformations allow the programmer to write a pair
of functions that transforms the \XXX[TODO]

>>>>

Second, this well-defined internal representation allows Nail to generate the external format from
the internal representation, without requiring the programmer to write additional code.

@@

Third, Nail introduces two abstractions, \emph{dependent fields} and \emph{transformations}, to
elegantly handle problematic structures, such as offset fields or checksums. Dependent fields are
fields in the protocol whose value depends in some way on the value or layout of other parts of the
format. Transformations allow the programmer to escape the generated code to modify the raw data and
interact with dependent fields in a controlled manner.
